Bayesian Parameter Inference for Nonlinear Stochastic Differential Equation Models

### 2015-08-13 TeXFiles/Manuscript_Albert_Ulzega.tex
paper with intro etc.

\title[Parameter inference with stochastic models]{Parameter Inference for a Simple Stochastic Hydrological Model}

\author[cor1]{Carlo Albert}%$^{1,2}$
\address{Eawag, Swiss Federal Institute of Aquatic Science and Technology, 8600 Dübendorf, Switzerland}
%\address{$^2$Address Two, Neverland}
\ead{carlo.albert@eawag.ch}

\author{Simone Ulzega}
\address{Eawag, Swiss Federal Institute of Aquatic Science and Technology, 8600 Dübendorf, Switzerland}
\ead{simone.ulzega@eawag.ch}

@@

\begin{abstract}
\textbf{[TO BE WRITTEN]} This document describes the  preparation of an article using \LaTeXe\ and \verb"iopart.cls" (the IOP \LaTeXe\ preprint class file).
This class file is designed to help
authors produce preprints in a form suitable for submission to any of the
journals published by IOP Publishing.
Authors submitting to any IOP journal, i.e.\
both single- and double-column ones, should follow the guidelines set out here.
On acceptance, their TeX code will be converted to
the appropriate format for the journal concerned.

@@

%Uncomment for PACS numbers title message
\pacs{00.00, 20.00, 42.10}
% Keywords required only for MST, PB, PMB, PM, JOA, JOB?
\vspace{2pc}
\noindent{\it Keywords}: Article preparation, IOP journals
% Uncomment for Submitted to journal title message
\submitto{\NJP}
% Comment out if separate title page not required
%\maketitle

@@

In this paper, we study the feasibility of a full Bayesian inference, for simple conceptual rainfall-runoff models with scale-invariant noise.
We start with a single linear reservoir and describe the dynamics of its water content, $S(t)$, by means of the stochastic differential equation
\begin{equation}\label{sde}
\dot{S}(t) = r(t) - \frac{1}{K}\left(1+\frac{\gamma}{2}\right) S(t)
+
\sqrt{\frac{\gamma}{K}} S(t){\eta}(t)\,,
\end{equation}
where $r(t)$ describes the rain input and $\eta(t)$ denotes white noise, i.e.,
\begin{equation}\label{whitenoise}
\langle\eta(t)\eta(t')\rangle = \delta(t-t')\,.
\end{equation}
Due to the state-dependence of the noise term, Eq.~(\ref{sde}) is ill-defined unless we specify the discretization convention.
For computational reasons, we work with the {\em Stratonovich} convention.
Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{\eq}=Kr_0$ (see below).

The scale-invariance of the noise term renders the noise parameter $\gamma$ dimensionless.
Before we proceed, we replace the state variable $S(t)$ and parameter $K$ by dimensionless quantities $q(t)$ and $\beta$, respectively, by means of the transformations
\begin{equation}
  \beta=\sqrt{\frac{T\gamma}{K}}\,,\quad
  S(t)=\frac{T\gamma r(t)}{\beta^2}e^{\beta q(t)}\,.
\end{equation}
W.r.t. these new  variables and parameters, the model equation (\ref{sde}) becomes
\begin{equation}
  \dot q(t)
  =
  \frac{\beta}{T\gamma}e^{-\beta q(t)}
  -
  \frac{1}{T}\rho(t)
  +
  \frac{1}{\sqrt{T}}\eta(t)\,,
\end{equation}
with
\begin{equation}
  \rho(t)
  =
  \frac{T}{\beta}\frac{d}{dt}\ln r(t)
  +
  \frac{(2+\gamma)\beta}{2\gamma}\,.
\end{equation}

According to Ref.~\cite{lau_2007}, the probability $P(q_1,T|q_0,0)$ of finding the system in a state $q$ at time $t = T$ given that it was in an initial state $q_0$ at time $t = 0$, can be expressed in the form of a path integral as
\begin{equation}\label{pathint}
P(q_1,T|q_0,0)
=
\frac{1}{Z}
\int
e^{-{\mathcal S}[q,\dot q]}
\delta(q(T)-q_1)
\delta(q(0)-q_0)
\mathcal{D}q \,,
\end{equation}
where the integral extends over all paths $q:[0,T]\rightarrow \mathbb R$.
The path-measure $\mathcal Dq$ is formally written as an infinite product
\begin{equation}\label{pathmeasure_q}
{\mathcal Dq}
=
\prod_{t}
dq(t)\,.
\end{equation}
The {\em action} is a functional on the space of paths and reads as
\begin{equation}\label{action1}
{\mathcal S}[{q},\dot q]
=
\frac{1}{T}
\int_0^T \dt \left\{
\frac{1}{2}
\left(
    T\dot q(t)
    +
    \rho(t)
    -
    \frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2
    -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
\right\} \,.
\end{equation}
Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $q(t)$.

We introduce the time-dependent {\em Hamiltonian}
\begin{equation}\label{H}
  \mathcal{H}(q,t)
  =
  \frac{1}{\gamma}e^{-\beta q}+q\rho(t)\,,
\end{equation}
and rewrite the action~(\ref{action1}) as,
\begin{eqnarray}\label{action}
{\mathcal S}[{q},\dot q] \nonumber
\\
= \frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    T\frac{\partial \mathcal{H}(q,t)}{\partial t} -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
     %\frac{(2+\gamma)\beta^2}{4\gamma}
\right\} \nonumber
\\
+ \mathcal{H}(q(T),T) - \mathcal{H}(q(0),0) \nonumber
\\
= \frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    Tq(t)\dot\rho(t) -
     \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
%    \frac{(2+\gamma)\beta^2}{4\gamma}
\right\}  \nonumber
\\
+
    \frac{1}{\gamma}e^{-\beta q(T)}+q(T)\rho(T)
   -\frac{1}{\gamma}e^{-\beta q(0)}-q(0)\rho(0)
\,.
\end{eqnarray}

Next, we calculate the equilibrium distribution $P_{\eq}(q) = \lim_{T\rightarrow\infty} P(q,T|q_0,0)$ in the simple case of a constant rain input $r(t)\equiv r_{0}$. Systems at thermal equilibrium must fulfill detailed balance, i.e.,
\begin{equation}\label{detailed_balance}
P(q_1 t_1 | q_0 t_0 ) P_{\eq}(q_0) = P(q_0 t_1 | q_1 t_0 ) P_{\eq}(q_1) \,.
\end{equation}
After plugging in (\ref{pathint}) and (\ref{action}), and using the transformation $q(t) \rightarrow q(-t)$, which maps paths from $q_0$ to $q_1$ to paths from $q_1$ to $q_0$, we get, since $\dot\rho(t)= 0$,
\begin{equation}\label{Peq}
  P_{\eq}(q)
  \propto
  e^{-2\mathcal{H}(q)}\,.
\end{equation}
Transforming back to the original variables, it turns out that $P_{\eq}(S)$ is described by an inverse gamma distribution with scale parameter $2Kr_{0}/\gamma$ and shape parameter $(2+\gamma)/\gamma$, i.e.,
\begin{equation}\label{inverse_gamma}
  P_{\eq}(S)
  \propto
  S^{-2(1+\gamma)/\gamma}e^{-2Kr_{0}/(\gamma S)}\,,
\end{equation}
whose mean equals the equilibrium solution of the unperturbed system ($\gamma=0$),
\begin{equation}
  \langle S\rangle_{\eq}=Kr_{0},
\end{equation}
and whose variance, for $\gamma< 2$, is given by
\begin{equation}
  \langle (S - \langle S\rangle_{\eq})^2\rangle_{\eq}
  =
  K^2r_{0}^2
  \frac{\gamma}{2-\gamma}\,.
\end{equation}
Note that the variance diverges, for $\gamma\geq 2$.
The power-law decay of the inverse gamma distribution is reminiscent of the scale-invariance of the error model.

Let us now consider two time-series of {\em observed} rain inputs and discharge outputs, $r_s=r(t_s)$ and $y_s=S(t_s)/K$, respectively, measured at times $0=t_1<t_2<\dots < t_{n+1}=T$.
For the measurement error model, we simply set
\begin{equation}\label{data}
  \ln\frac{y_s}{r_s}
  =
  \beta q_s
  +
  \sigma\epsilon_s\,,\quad s=1,\dots,n+1\,,
\end{equation}
where $\epsilon_s$ are uncorrelated standard normal random variables.

The goal is to infer the posterior parameter distribution, for the parameters $\bt=(\beta,\gamma)^T$, based on the measured time series $\vc r$ and $\vc y$.
For the time being we assume constant priors.
Then, the posterior reads as
\begin{equation}\label{posterior_pathint}
  f(\bt | \vc y, \vc r)
  \propto
  \int
  \exp\bigg[
    -{{\mathcal S}}[{q},\dot q]
    -\frac{1}{2}
    \sum_{s=1}^{n+1}
    \frac
    {(\ln(y_s/r_s)-{\beta q_s})^2}
    {\sigma^2}
    %-\ln(K\gamma)
  \bigg]
  {\mathcal Dq}
  \,.
\end{equation}
Since the integral on the r.h.s. of (\ref{posterior_pathint}) is very expensive to calculate, we will sample, simultaneously, parameter vectors $\bt$ and discretized system realizations $q(t)$ directly from an appropriate discretization of the kernel of (\ref{posterior_pathint}).
Since the associated distribution is very high dimensional and strongly correlated, we employ the method called {\em Hamiltonian Monte Carlo} (HMC) introduced by Ref.~\cite{duane_1987}.
This method interprets the exponent of the kernel of the posterior as the potential of a one-dimensional statistical mechanical system. Each degree of freedom, $q(t)$ and $\bt$ in our case, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$, respectively, and the system is defined by the  Hamiltonian
\begin{equation}\label{Hamiltonian}
    \mathcal{H}_{\mbox{\tiny HMC}}(q,\bt; p,{\pmb\pi})
    =
    K( p,{\pmb\pi}) + V( q,\bt)\,,
\end{equation}
with
\begin{equation}\label{K}
   K( p,{\pmb\pi})
   =
   \int_0^T \frac{ p^2(t)}{2m(t)}dt
   + \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,
\end{equation}
where $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
The posterior (\ref{posterior_pathint}) is expressed by the phase space path integral
\begin{equation}\label{phaseSpacePathInt}
    f(\bt | \vc y, \vc r)
  \propto
  \int
  e^{-\mathcal {H}_{\mbox{\tiny HMC}}(q,\bt; p,{\pmb\pi})}
  {\mathcal Dp}
  \wedge
  {\mathcal Dq}
  \wedge
  d{\pmb\pi}
  \,.
\end{equation}

The HMC method, which is a combination of the {\em Metropolis algorithm} of Ref.~\cite{metropolis_1953} and {\em molecular dynamics} methods \textbf{[ref?]}, iterates the following steps,
\begin{enumerate}
  \item
  Momenta are sampled from the Gaussian distribution defined by (\ref{K}).
  \item
  A volume-preserving and time-reversible solution to a discretization of Hamilton's equations is calculated numerically.
  \item
  The discretization error on the energy preservation due to the previous step is corrected by a Metropolis acceptance/rejection step.
\end{enumerate}
The last step is the standard Metropolis algorithm, while the first two steps allow us to make relatively large jumps in phase space while maintaining a relatively large acceptance rate.

Let us first write down the discretization of the path-integral (\ref{phaseSpacePathInt}).
Therefore, let us assume that the measurement time points $\left\{ y_s \right\}_{s=1,\dots, n+1}$ of the time series (\ref{data}) are equidistantly distributed on the time interval $[0,T]$, with $t_1=0$ and $t_{n+1}=T$.
Each interval between two consecutive data points is further partitioned into $j$ bins, such that we have a total of $nj+1=N>>1$ discretization points.
Path-integral (\ref{posterior_pathint}) is approximated by an ordinary integral, with the approximate path-measure
\begin{equation}
  \mathcal Dp\wedge\mathcal Dq
  \approx
  \prod_i dp_i\wedge dq_i
\end{equation}
and the discretized versions of $K( p,{\pmb\pi})$ and $V( q,\bt)$
\begin{eqnarray}
   K( p,{\pmb\pi}) \approx
   \sum_{i=1}^N
   \frac{ p_i^2}{2m_q}\Delta t
   +
   \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,\label{Kdisc}
   \\
   V(q,\bt) \approx \frac{\Delta t}{T} \sum_{i=2}^{N}
   \left\{ \frac{1}{2} T^2 \dot q_i^2 + \frac{1}{2}
     \left( \rho_i-\frac{\beta}{\gamma}e^{-\beta q_i} \right)^2 -
    \frac{\beta^2}{2\gamma} e^{-\beta q_i} - T q_i\dot\rho_i \right\}  \nonumber
  \\
  +
  \frac{1}{\gamma}
  e^{-\beta q_N}
  +
  q_N \rho_{N}
  -
  \frac{1}{\gamma}
  e^{-\beta q_1}
  -
  q_1 \rho_{1}
  \nonumber
  \\
  +
  \sum_{s=1}^{n+1}
  \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta q_{(s-1)j+1}})^2}{2\sigma^2}
  \label{Vdisc}
   \,,
\end{eqnarray}
with
\begin{equation}
  \dot q_i = \frac{q_i-q_{i-1}}{\Delta t}\,,
  %\quad
%  \bar q_i = \frac{q_i+q_{i-1}}{2}\,,
\end{equation}
and
\begin{equation}\label{rhodisc}
\rho_i = \frac{T}{\beta} \frac{\ln(r_{i}/r_{i-1})}{\Delta t}
+
\frac{(2+\gamma)\beta}{2\gamma}
\,,\quad
\dot\rho_i = \frac{\rho_i-\rho_{i-1}}{\Delta t}\,.
\end{equation}
Note that we've neglected terms of order $\mathcal O(N^{-1/2})$ in the action (\ref{Vdisc}) and that we've assumed $r_i$ to be the discretization of an at least twice differentiable interpolation of the input measurements.

For strongly varying rain input or high-frequency output data, $\Delta t$ needs to be chosen small. Thus, the harmonic part of the Hamiltonian becomes stiff and, consequently, the time-step of MD needs to be chosen small, which leads to slow convergence.
To solve this problem, we change coordinates, such that the harmonic part of the Hamiltonian becomes (partly) diagonal and the particle interaction moves to the potential.
One option would be to use the Fourier modes of ${\eta}(t)$ as configurational variables. Calculating the potential requires then an inversion of a Toeplitz matrix. This approach is interesting because, in our case, it allows for an approximate analytical integration of high-frequency modes as outlined in Ref.~\cite{doll_1985}.

We choose however another approach, the so-called \emph{staging} as described in Ref.~\cite{tuckerman_1992}, and diagonalize the harmonic part only in between the measurement points.
We first note that the Hamiltonian (\ref{Hamiltonian}) can be seen as the classical Hamiltonian of a polymer chain with harmonic bonds between neighbouring beads, described by the first harmonic term of $V(q,\bt)$, in an external field described by the other non-harmonic terms of $V(q,\bt)$.
Furthermore, the model parameters are viewed as additional degrees of freedom that couple with all other degrees of freedom.
Sampling system realizations, each defined by a set of coordinates $\left\{ q_i\right\}$ and parameters $\bt$, is equivalent to sampling states in the polymer dynamics.

>>>>

\title[Parameter inference with stochastic models]{Bayesian Parameter Inference for Nonlinear Stochastic Differential Equation Models}

\author[cor1]{Carlo Albert}%$^{1,2}$
\address{Eawag, Swiss Federal Institute of Aquatic Science and Technology, 8600 D\"ubendorf, Switzerland}
%\address{$^2$Address Two, Neverland}
\ead{carlo.albert@eawag.ch}

\author{Simone Ulzega}
\address{Eawag, Swiss Federal Institute of Aquatic Science and Technology, 8600 D\"ubendorf, Switzerland}
\ead{simone.ulzega@eawag.ch}

@@

\begin{abstract}
Bayesian statistics has become an indispensable tool in many applied sciences, for the purpose of uncertainty analysis.  
Inferring parametric uncertainty, for stochastic differential equation models, however, is a computationally hard problem due to the high dimensional integrals that have to be calculated.
Here, we consider the generic problem of calibrating a 1D SDE model to time series and quantifying the ensuing parametric uncertainty.
We re-interpret the Bayesian posterior distribution, for model parameters, as the partition function of a statistical mechanical system and employ a Hamiltonian Monte Carlo algorithm to calculate it.
Depending on the number of discretization points and the number of measurement points the dynamics of this system happens on very different time scales.
Thus, we employ a multiple time scale integration together with a suitable re-parametrization to derive an efficient inference algorithm.
While the algorithm is presented by means of a simple SDE model from hydrology, it is readily applicable to a wide range of inference problems.
Furthermore the algorithm is highly parallelizable.

@@

%Uncomment for PACS numbers title message
\pacs{00.00, 20.00, 42.10}
% Keywords required only for MST, PB, PMB, PM, JOA, JOB?
\vspace{2pc}
\noindent{\it Keywords}: Bayesian parameter inference, Hamiltonian Monte Carlo, time series analysis, stochastic differential equations

% Uncomment for Submitted to journal title message
\submitto{\NJP}
% Comment out if separate title page not required
%\maketitle

@@

In many applied areas of research phenomenological models are employed to predict the behaviour of complex systems of all sorts.
Parameters of these models often don't have a direct physical meaning and need to be inferred from measurements.
In order to make reliable predictions with such models it is important to describe the dominant errors in the model and to quantify the parametric uncertainty resulting from the inference process.

{\em Bayesian statistics} describes knowledge about parameters through probability distributions and model-based learning through a consistent update rule. It is thus well suited to quantify parametric uncertainty resulting from a model based inference process.
Bayesian statistics is commonly used in many applied sciences and grows in importance in the physics community as well \cite{vonToussaint2011bayesian}.

A faithful description of the dominant errors in a model naturally leads to {\em stochastic differential equations} (SDEs).
The kind of problems we consider here are the calibration of ordinary 1D SDE models to noisy time series and the quantification of the resulting parametric uncertainty.
Whilst the techniques we use are generic they will be presented by means of a simple yet non trivial SDE model from hydrology.

Bayesian parameter inference for SDE models is computationally very expensive, as the posterior probability density for the parameters is a {\em path-integral}.
Problems of this kind are commonly solved by means of {\em Monte Carlo} methods that are based on simulating model realizations and comparing them to the data. Algorithms of this kind are {\em particle filters} \cite{chopin_2013_SMC2}. For the case of linear SDEs that are coupled to nonlinear deterministic ODEs, more efficient algorithms of this kind can be derived \cite{tomassini_2009_smoothing, reichert_2009_timedepParameters}.
The problem with these simulation-based methods is their inefficiency in the presence of many data points.
One solution is to map the output space to a smaller dimensional space of {\em summary statistics}, and acccept/reject proposed model parameters depending on how compatible associated model runs are with the data in terms of these summary statistics.
Such {\em Approximate Bayes Computations} \cite{Albert_2013_ABC} are relatively easy to apply as they only require us to run the simulator. However, it is largely an unsolved problem how to choose the summary statistics so as to achieve a sufficient approximation of the posterior parameter distribution.

Exact inference algorithms of high efficiency can be derived from a reinterpretation of the posterior distribution as the partition function of a statistical mechanical system and simulation of the dynamics of the latter.
Such {\em Hamiltonian Monte Carlo} (HMC) algorithms \cite{duane_1987_HMC} incorporate the data points for the suggestion of new parameters whereby much higher acceptance rates are achieved.
The drawback of these methods is that the model equations need to be known and derivatives have to be calculated. The latter problem, however, is largely remedied by the use of automated differentiation algorithms.

Tuning of HMC algorithms is a non-trivial matter.
Two sorts of tuning parameters need to be adjusted: (i) those of the {\em kinetic energy} of the statistical mechanical system and (ii) those of the numerical integration scheme of Hamilton's equation in the molecular dynamics part of the HMC algorithm.
It has been shown that efficiency is gained when the kinetic term is made dependent of the configuration of the statistical mechanical system \cite{girolami_2011_HMC}.
Here, we explore a different, computationally simper route.
Depending on the number of discretization points needed to approximate the original SDE system and the number of measurement points, the dynamics of the statistical mechanical system happens on very different time scales.
Thus, we employ a multiple time scale integration technique, for the simulation of the statistical mechanical system \cite{tuckerman1993efficient}.
At lest for 1D SDEs we always find a parametrization, which allows to partly uncouple the dominant harmonic part of the Hamiltonian in the form of harmonic oscillators and integrate them analytically.

The resulting algorithm is very efficient and, through the use of automated differentiation, easily applicable to a wide range of inference problems with SDEs.
Furthermore, it is easily parallelizable.

\section{An Exemplary SDE Model}

The methods we present are generic and applicable to a wide range of SDE models.
However, with later applications in mind and to make things concrete, we present them by means of a simple yet non-trivial model from hydrology.
Consider a hydrological catchment whose dynamics at the observation time scale is well described by a linear reservoir and whose other processes happen at much shorter time scales so that they can be described by white noise.
Furthermore, assume this noise to scale linearly with the system state, $S(t)$, which is the water content in the reservoir.
The model equation is thus given by the SDE
\begin{equation}\label{sde}
\dot{S}(t) = r(t) - \frac{1}{K}\left(1+\frac{\gamma}{2}\right) S(t)
+
\sqrt{\frac{\gamma}{K}} S(t){\eta}(t)\,,
\end{equation}
where $r(t)$ denotes the time varying rain input and $\eta(t)$ denotes white noise, i.e.,
\begin{equation}\label{whitenoise}
\langle\eta(t)\eta(t')\rangle = \delta(t-t')\,.
\end{equation}
Eq.~(\ref{sde}) is to be understood in the {\em Stratonovich} sense.
Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{\eq}=Kr_0$ (see eq. (\ref{equilibrium_mean}) below).
Scale-invariance of the error term in (\ref{sde}) leads to power law tails in the system's distribution (see eq. (\ref{inverse_gamma})), which is in line with the observation that errors in hydrological models are often fat-tailed (ref).

Conceptual models of this kind are commonly used in hydrology, for the purpose of predicting rainfall-runoff behaviour, for natural and urban catchments (see, e.g., \cite{breinholt_2011_SDE}).
Furthermore, by means of the transformation $S(t)=1/n(t)$, eq. (\ref{sde}) turns into a model that has been suggested as a phenomenological description of the dynamics of the neutron density in nuclear reactors and extensively studied \cite{dutre_1977_SDE, fujisaka_1986_intermittency}.

Here, we consider the input $r(t)$ to be a smooth and nowhere vanishing function. Realistic multi-fractal rain inputs will be studied elsewhere.
Indeed, the focus of this work is not on hydrological modeling, but on generic methods for parameter inference with SDE models that are calibrated to observed time-series.
Here, we assume the observed time-series, $y_s$, to be the outflow of the reservoir, $S(t)/K$, which is observed, at times $0=t_1<t_2<\dots < t_{n+1}=T$, with multiplicative independent log-normal errors, i.e.,
\begin{equation}\label{data}
  \ln y_s
  =
  \ln\frac{S(t_s)}{K}
  +
  \sigma\epsilon_s\,,\quad s=1,\dots,n+1\,.
\end{equation}
For simplicity, we assume $\sigma$ as well as the input $r(t)$ to be known so that we are left with the task of inferring parameter combinations $(K,\gamma)$ that are compatible with the data (\ref{data}).
Here, "compatible" is meant in the Bayesian sense, in which knowledge about parameters is expressed in terms of a probability distribution.
In general, prior knowledge about parameters, $\bt$, is expressed in terms of a prior probability distribution, $f_{prior}(\bt)$, and the posterior, which combines prior knowledge with knowledge gained from data, is calculated by means of Bayes' formula
\begin{equation}\label{Bayes}
  f_{post}(\bt|\vc y)
  =
  \frac{f_{prior}(\bt)L(\vc y|\bt)}
  {f_{prior}(\bt')L(\vc y|\bt')d\bt'}\,,
\end{equation}
where $L(\vc y|\bt)$ is the so-called {\em likelihood function}, that is, the probability distribution, for model outputs given model parameters, evaluated at the measured data.
Here, we assume prior knowledge to be that parameters must be non-negative, but otherwise we assume the prior to be flat on an area large enough to encompass the area where the likelihood function is significantly different from zero. That is, the posterior is data-driven.

Before we set out to derive the likelihood function from the model equations (\ref{sde}) through (\ref{data}) let us express the parameters and state variable with dimensionless quantities.
The noise parameter $\gamma$ is already dimensionless due to scale-invariance of the noise term.
We replace the state variable $S(t)$ and parameter $K$ by dimensionless quantities $q(t)$ and $\beta$, respectively, by means of the transformations
\begin{equation}
  \beta=\sqrt{\frac{T\gamma}{K}}\,,\quad
  S(t)=\frac{T\gamma r(t)}{\beta^2}e^{\beta q(t)}\,.
\end{equation}
W.r.t. these new  variables and parameters, the model equation (\ref{sde}) becomes
\begin{equation}
  \dot q(t)
  =
  \frac{\beta}{T\gamma}e^{-\beta q(t)}
  -
  \frac{1}{T}\rho(t)
  +
  \frac{1}{\sqrt{T}}\eta(t)\,,
\end{equation}
with
\begin{equation}
  \rho(t)
  =
  \frac{T}{\beta}\frac{d}{dt}\ln r(t)
  +
  \frac{(2+\gamma)\beta}{2\gamma}\,.
\end{equation}
For our algorithm it is important to have the model equation in a form where the noise term does not depend on any of the variables that have to be inferred, i.e., on neither the state variable nor the parameters that have to be inferred.
At least in 1D this can always be achieved through re-parametrization.

The probability $P(q_1,T|q_0,0)$ of finding the system in a state $q_1$, at time $t = T$ given that it was in an initial state $q_0$ at time $t = 0$, is expressed in the form of a {\em path-integral} as
\begin{equation}\label{pathint}
P(q_1,T|q_0,0)
=
\frac{1}{Z}
\int
e^{-{\mathcal S}[q,\dot q]}
\delta(q(T)-q_1)
\delta(q(0)-q_0)
\mathcal{D}q \,,
\end{equation}
where the integral extends over all paths $q:[0,T]\rightarrow \mathbb R$.
The path-measure $\mathcal Dq$ is formally written as the infinite product
\begin{equation}\label{pathmeasure_q}
{\mathcal Dq}
=
\prod_{t}
dq(t)\,.
\end{equation}
The {\em action} is a functional on the space of paths and reads as
\begin{equation}\label{action1}
{\mathcal S}[{q},\dot q]
=
\frac{1}{T}
\int_0^T \dt \left\{
\frac{1}{2}
\left(
    T\dot q(t)
    +
    \rho(t)
    -
    \frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2
    -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
\right\} \,.
\end{equation}
Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $q(t)$ \cite{lau_2007_StateDepDiff}.

We introduce the time-dependent {\em Hamiltonian}
\begin{equation}\label{H}
  \mathcal{H}(q,t)
  =
  \frac{1}{\gamma}e^{-\beta q}+q\rho(t)\,,
\end{equation}
and rewrite action~(\ref{action1}) as,
\begin{eqnarray}\label{action}
{\mathcal S}[{q},\dot q] \nonumber
\\
= \frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    T\frac{\partial \mathcal{H}(q,t)}{\partial t} -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
     %\frac{(2+\gamma)\beta^2}{4\gamma}
\right\} \nonumber
\\
+ \mathcal{H}(q(T),T) - \mathcal{H}(q(0),0) \nonumber
\\
= \frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    Tq(t)\dot\rho(t) -
     \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
%    \frac{(2+\gamma)\beta^2}{4\gamma}
\right\}  \nonumber
\\
+
    \frac{1}{\gamma}e^{-\beta q(T)}+q(T)\rho(T)
   -\frac{1}{\gamma}e^{-\beta q(0)}-q(0)\rho(0)
\,.
\end{eqnarray}

Properties of (a transformed version of) (\ref{sde}) have been derived in \cite{} and we do not repeat them here, as the focus of this paper is not on the properties of this particular model. However, in order to prove two claims made earlier, we calculate the {\em equilibrium distribution} $P_{\eq}(q) = \lim_{T\rightarrow\infty} P(q,T|q_0,0)$ in the simple case of a constant input $r(t)\equiv r_{0}$. After plugging (\ref{pathint}) and (\ref{action}) into the detailed balance condition,
\begin{equation}\label{detailed_balance}
P(q_1 t_1 | q_0 t_0 ) P_{\eq}(q_0) = P(q_0 t_1 | q_1 t_0 ) P_{\eq}(q_1) \,,
\end{equation}
and using transformation $q(t) \rightarrow q(-t)$, which maps paths from $q_0$ to $q_1$ onto paths from $q_1$ to $q_0$, we get, since $\dot\rho(t)= 0$,
\begin{equation}\label{Peq}
  P_{\eq}(q)
  \propto
  e^{-2\mathcal{H}(q)}\,.
\end{equation}
Transforming back to the original variables, it turns out that $P_{\eq}(S)$ is an inverse gamma distribution with scale parameter $2Kr_{0}/\gamma$ and shape parameter $(2+\gamma)/\gamma$, i.e.,
\begin{equation}\label{inverse_gamma}
  P_{\eq}(S)
  \propto
  S^{-2(1+\gamma)/\gamma}e^{-2Kr_{0}/(\gamma S)}\,,
\end{equation}
whose mean equals the equilibrium solution of the unperturbed system ($\gamma=0$),
\begin{equation}\label{equilibrium_mean}
  \langle S\rangle_{\eq}=Kr_{0},
\end{equation}
and whose variance, for $\gamma< 2$, is given by
\begin{equation}
  \langle (S - \langle S\rangle_{\eq})^2\rangle_{\eq}
  =
  K^2r_{0}^2
  \frac{\gamma}{2-\gamma}\,
\end{equation}
and diverges, for $\gamma\geq 2$.
The power-law decay of the inverse gamma distribution is reminiscent of the scale-invariance of the error model.

If we denote the parameter vector $\bt=(\beta,\gamma)^T$ and assume a flat prior as described above, the posterior (\ref{Bayes}), as a function of $\bt$, is proportional to likelihood function, which is expressed as a path-integral as follows
\begin{equation}\label{posterior_pathint}
  f(\bt | \vc y)
  \propto
  \int
  \exp\bigg[
    -{{\mathcal S}}[{q},\dot q]
    -\frac{1}{2}
    \sum_{s=1}^{n+1}
    \frac
    {(\ln(y_s/r(t_s))-{\beta q(t_s)})^2}
    {\sigma^2}
    %-\ln(K\gamma)
  \bigg]
  {\mathcal Dq}
  \,.
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Inference Algorithm}

In order to derive an efficient algorithm to draw parameter samples from (\ref{posterior_pathint}) we interpret it as the partition function of a 1D statistical mechanical system and simulate the dynamics of the latter, employing the so-called {\em Hamiltonian Monte Carlo} (HMC) algorithm \cite{duane_1987_HMC}.
The model parameters $\bt$ are interpreted as additional dynamical degrees of freedom coupling to the field $q(t)$.
Each degree of freedom, $q(t)$ and $\bt$ in our case, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$, respectively, and the system is defined by the  Hamiltonian
\begin{equation}\label{Hamiltonian}
    \mathcal{H}_{\mbox{\tiny HMC}}(q,\bt; p,{\pmb\pi})
    =
    K( p,{\pmb\pi}) + V( q,\bt)\,,
\end{equation}
where
\begin{equation}\label{K}
   K( p,{\pmb\pi})
   =
   \int_0^T \frac{ p^2(t)}{2m(t)}dt
   + \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,
\end{equation}
and $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
The posterior (\ref{posterior_pathint}) is expressed by the phase space path integral
\begin{equation}\label{phaseSpacePathInt}
    f(\bt | \vc y)
  \propto
  \int
  e^{-\mathcal {H}_{\mbox{\tiny HMC}}(q,\bt; p,{\pmb\pi})}
  {\mathcal Dp}
   {\mathcal Dq}
   d{\pmb\pi}
  \,.
\end{equation}

The HMC method, which is a combination of the {\em Metropolis algorithm} \cite{metropolis_1953_MRT2} and {\em molecular dynamics} methods \cite{Alder_1959_MD, Rahman_1964_MD}, iterates the following steps:
\begin{enumerate}
  \item
  Momenta are sampled from the Gaussian distribution defined by (\ref{K}).
  \item
  The system is integrated by means of a volume-preserving and time-reversible solution of a discretization of Hamilton's equations.
  \item
  The discretization error on the energy preservation due to the previous step is corrected by a Metropolis acceptance/rejection step.
\end{enumerate}
The last step is the standard Metropolis algorithm, while the first two steps allow us to make arbitrarily large jumps in phase space while maintaining an arbitrarily large acceptance rate.
